---
title: "Chad's 2025 MLB Report"
output: github_document
knit: (function(input, ...) { rmarkdown::render(input, output_file = "README.md", envir = globalenv()) })
---

*Interested in the underlying code that builds this report?* Check it out on GitHub: [mlb25](https://github.com/chadallison/mlb25){target="_blank"}

___

### Contents

- [Team Standings]
- [Runs Scored and Allowed per Game]
- [NPR]
- [Offensive and Defensive NPR]
- [Run Differentials]
- [Pythagorean Wins]
- [Actual vs Pythagorean Win Percentages]
- [Winning and Losing Streaks]
- [Home and Away Splits]
- [Rolling Window Pythagorean Wins]
- [Rolling Window NPR]
- [Records Against Teams with Winning vs Losing Records]
- [Performance Against Teams with Winning Record]
- [NPR Trends]
- [Runs Scored and Volatility]
- [Performance by Strength of Schedule]
- [One Run Games]
- [Pythagorean Wins in Last Ten]

___

```{r fold_code, include = F}
knitr::knit_hooks$set(source = function(x, options) {
    hook.r = function(x, options) {
      fence = "```"
      language = tolower(options$engine)
      if (language == "node") language = "javascript"
      if (!options$highlight) language = "text"
      if (!is.null(options$fold_code)) {
        paste0("\n\n", "<details><summary>View Code</summary>\n", fence, language,
               "\n", x, fence, "\n\n", "</details>\n")
      } else paste0('\n\n', fence, language, '\n', x, fence,  '\n\n')
    }
    x = knitr:::hilight_source(x, "markdown", options)
    hook.r(paste(c(x, ""), collapse = "\n"), options)
})
```

```{r message = F, warning = F, include = F, fold_code = T}
library(tidyverse)
library(tvthemes)
library(janitor)
library(glue)
library(zoo)
library(baseballr)

theme_custom = theme_avatar() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 9, vjust = 2.5, face = "italic"),
        plot.caption = element_text(face = "italic"),
        panel.grid.major = element_line(linewidth = 0.5, colour = "#DFDAD1"),
        panel.grid.minor = element_line(linewidth = 0.5, colour = "#DFDAD1"))

theme_set(theme_custom)
```

```{r message = F, warning = F, include = F}
end_games_raw = mlb_schedule(season = 2025, level_ids = "1")

end_games = end_games_raw |>
  filter(series_description == "Regular Season" & status_detailed_state == "Final") |>
  select(date, game_pk, away_score = teams_away_score, away_team = teams_away_team_name,
         home_team = teams_home_team_name, home_score = teams_home_score) |>
  filter(!is.na(away_score) & !is.na(home_score)) |>
  mutate(win_team = ifelse(home_score > away_score, home_team, away_team),
         lose_team = ifelse(home_score > away_score, away_team, home_team),
         win_score = ifelse(home_score > away_score, home_score, away_score),
         lose_score = ifelse(home_score > away_score, away_score, home_score))

# writing and reading prevents baseballr message
write_csv(end_games, "data/end_games.csv")
end_games = read_csv("data/end_games.csv", show_col_types = F)

max_margin = end_games |>
  mutate(margin = win_score - lose_score,
         margin_pct = percent_rank(margin)) |>
  filter(margin_pct > 0.95) |>
  pull(margin) |>
  min()

# change this to true to truncate top 5% of win margins
truncate_blowouts = FALSE

if (truncate_blowouts) {
  end_games = end_games |> filter(win_score - lose_score <= max_margin)
}
```

```{r include = F}
all_teams = sort(unique(c(end_games$home_team, end_games$away_team)))

team_abbrevs =  c(
  "ARI", "ATH", "ATL", "BAL", "BOS", "CHC",
  "CWS", "CIN", "CLE", "COL", "DET", "HOU",
  "KC", "LAA", "LAD", "MIA", "MIL", "MIN",
  "NYM", "NYY", "PHI", "PIT", "SD", "SF",
  "SEA", "STL", "TB", "TEX", "TOR", "WSH"
)

team_hex = c(
  "#A71930", "#003831", "#CE1141", "#DF4601", "#BD3039", "#0E3386",
  "#27251F", "#C6011F", "#0C2340", "#33006F", "#FA4616", "#EB6E1F",
  "#004687", "#BA0021", "#005A9C", "#00A3E0", "#FFC52F", "#D31145",
  "#002D72", "#003087", "#E81828", "#FDB827", "#2F241D", "#FD5A1E",
  "#0C2C56", "#C41E3A", "#8FBCE6", "#C0111F", "#134A8E", "#dcbdd0"
)

divisions <- c(
  "NL West", "AL West", "NL East", "AL East", "AL East", "NL Central",
  "AL Central", "NL Central", "AL Central", "NL West", "AL Central", "AL West",
  "AL Central", "AL West", "NL West", "NL East", "NL Central", "AL Central",
  "NL East", "AL East", "NL East", "NL Central", "NL West", "NL West",
  "AL West", "NL Central", "AL East", "AL West", "AL East", "NL East"
)

team_divisons = data.frame(team = all_teams, division = divisions) |>
  mutate(division = factor(division, levels = c("AL West", "AL Central", "AL East",
                                                "NL West", "NL Central", "NL East")))

teams_info = data.frame(team = all_teams, abb = team_abbrevs, hex = team_hex)

better_date = function(dt) {
  return(as.character(glue("{month(dt, label = T, abbr = F)} {day(dt)}, {year(dt)}")))
}

today_nice = better_date(Sys.Date())
```

### Team Standings

```{r echo = F}
team_wins = end_games |>
  count(win_team, name = "wins")

team_losses = end_games |>
  count(lose_team, name = "losses")

team_records = team_wins |>
  full_join(team_losses, by = c("win_team" = "lose_team")) |>
  mutate(wins = coalesce(wins, 0),
         losses = coalesce(losses, 0),
         gp = wins + losses,
         win_pct = round(wins / gp * 100, 2),
         record = as.character(glue("{wins}-{losses}"))) |>
  rename(team = win_team) |>
  inner_join(teams_info, by = "team")

team_records |>
  ggplot(aes(reorder(reorder(team, wins), win_pct), win_pct)) +
  geom_col(aes(fill = team), show.legend = F) +
  geom_text(aes(label = record), size = 3, hjust = -0.2) +
  coord_flip(ylim = c(0, max(team_records$win_pct) * 1.05)) +
  scale_fill_manual(values = team_hex) +
  labs(x = NULL, y = NULL, title = "2025 MLB Standings") +
  theme(axis.text.x = element_blank())
```

___

### Runs Scored and Allowed per Game

```{r echo = F}
home_scored_allowed = end_games |>
  select(date, team = home_team, scored = home_score, allowed = away_score) |>
  mutate(home_away = "home")

away_scored_allowed = end_games |>
  select(date, team = away_team, scored = away_score, allowed = home_score) |>
  mutate(home_away = "away")

team_scored_allowed = bind_rows(home_scored_allowed, away_scored_allowed) |>
  arrange(team, date)

team_rspg_rapg = team_scored_allowed |>
  group_by(team) |>
  summarise(rspg = sum(scored) / n(),
            rapg = sum(allowed) / n())

team_rspg_rapg |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(rspg, rapg)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = mean(team_rspg_rapg$rspg), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = mean(team_rspg_rapg$rapg), linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  labs(x = "Runs Scored per Game", y = "Runs Allowed Per Game",
       title = glue("Team Runs Scored/Allowed per Game as of {today_nice}")) +
  scale_x_continuous(breaks = seq(0, 15, by = 1)) +
  scale_y_continuous(breaks = seq(0, 15, by = 1))
```

___

### NPR

```{r echo = F}
end_with_npr = end_games |>
  inner_join(team_rspg_rapg, by = c("home_team" = "team")) |>
  rename(home_rspg = rspg, home_rapg = rapg) |>
  inner_join(team_rspg_rapg, by = c("away_team" = "team")) |>
  rename(away_rspg = rspg, away_rapg = rapg) |>
  mutate(home_exp = (home_rspg + away_rapg) / 2,
         away_exp = (away_rspg + home_rapg) / 2,
         home_off_npr = home_score - home_exp,
         home_def_npr = away_exp - away_score,
         away_off_npr = away_score - away_exp,
         away_def_npr = home_exp - home_score)

home_npr_by_game = end_with_npr |>
  transmute(date, team = home_team, off_npr = home_off_npr, def_npr = home_def_npr, home_away = "home")

away_npr_by_game = end_with_npr |>
  transmute(date, team = away_team, off_npr = away_off_npr, def_npr = away_def_npr, home_away = "away")

team_npr_by_game = bind_rows(home_npr_by_game, away_npr_by_game) |>
  arrange(team, date)

team_npr = team_npr_by_game |>
  group_by(team) |>
  summarise(off_npr = round(mean(off_npr), 3),
            def_npr = round(mean(def_npr), 3),
            ovr_npr = off_npr + def_npr)

team_npr |>
  mutate(pl = ifelse(ovr_npr >= 0, round(ovr_npr, 3), ""),
         nl = ifelse(ovr_npr < 0, round(ovr_npr, 3), "")) |>
  ggplot(aes(reorder(team, ovr_npr), ovr_npr)) +
  geom_col(aes(fill = team), show.legend = F) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  scale_fill_manual(values = team_hex) +
  scale_y_continuous(breaks = seq(-3, 3, by = 0.25)) +
  coord_flip(ylim = c(min(team_npr$ovr_npr) * 1.05, max(team_npr$ovr_npr) * 1.05)) +
  labs(x = NULL, y = "Overall NPR",
       title = glue("Team NPR standings as of {today_nice}"))
```

**What is NPR?**

NPR, Naive Performance Rating, is a metric I devised as a measure of team performance above/below expected. The logic behind it is this: I calculate each team’s expected runs scored in each game by taking the average of their runs scored per game and their opponent’s runs allowed per game. I then compare this expected value to the actual value of runs scored or allowed to calculate each team’s offensive and defensive NPR for each game. Here is an example.

Suppose the Cubs are playing the Cardinals. Let’s say the Cubs, on average, score 4.5 runs per game and allow 3.25 runs per game. And let’s say the Cardinals score 3.75 runs per game and allow 2.75 runs per game. We calculate the Cubs’ expected run value as the average of their runs scored per game and the Cardinals’ runs allowed per game, so (4.5 + 2.75) / 2 = 3.63. We would calculate the Cardinals’ expected run value the same way, so (3.75 + 3.25) / 2 = 3.5. We now have the Cubs’ expected run value as 3.63 and the Cardinals’ expected run value as 3.5.

Suppose that the final score of the game is a Cubs victory, 5-3. We would calculate the Cubs’ offensive NPR as their actual score minus their expected score: 5 - 3.63 = 1.37. We would calculate their defensive NPR as the Cardinals’ expected score minus their actual score: 3.5 - 3 = 0.5 (we do it in this order so positive values are good). For the Cardinals, their offensive NPR is their actual score minus their expected score, 3 - 3.5 = -0.5, and their defensive NPR is the Cubs’ expected score minus their actual score, 3.63 - 5 = -1.37. Notice how these numbers are opposite each other. So each team will have an offensive and defensive NPR for each game, which are aggregated in the plot above.

Of course, there are so many other factors that would play into a team’s true expected value, such as any injuries, starting pitchers, weather, and more. That is why I have named it Naive Performance Rating, because it assumes matchup metrics are independent of each other and does not take external factors into account. Which, of course, will lead to flaws in the metric, but is done for the sake of simplicity and interpretability.

___

### Offensive and Defensive NPR

```{r echo = F}
team_npr |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(off_npr, def_npr)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3) +
  scale_color_manual(values = team_hex) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(x = "Offensive NPR", y = "Defensive NPR",
       title = glue("Team offensive/defensive NPR as of {today_nice}")) +
  scale_x_continuous(breaks = seq(-3, 3, by = 0.25)) +
  scale_y_continuous(breaks = seq(-3, 3, by = 0.25))
```

___

### Run Differentials

```{r echo = F}
home_margins = end_games |>
  transmute(date, team = home_team, margin = home_score - away_score, home_away = "home")

away_margins = end_games |>
  transmute(date, team = away_team, margin = away_score - home_score, home_away = "away")

team_game_margins = bind_rows(home_margins, away_margins) |>
  arrange(team, date)

team_game_margins |>
  group_by(team) |>
  summarise(diff_per_game = sum(margin) / n()) |>
  inner_join(teams_info, by = "team") |>
  mutate(pl = ifelse(diff_per_game >= 0, round(diff_per_game, 2), ""),
         nl = ifelse(diff_per_game < 0, round(diff_per_game, 2), "")) |>
  ggplot(aes(reorder(abb, diff_per_game), diff_per_game)) +
  geom_col(aes(fill = team), show.legend = F) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  coord_flip() +
  scale_fill_manual(values = team_hex) +
  labs(x = NULL, y = "Avg. run differential",
       title = glue("Team avg. run differentials as of {today_nice}")) +
  scale_y_continuous(breaks = seq(-10, 10, by = 1))
```

___

### Pythagorean Wins

```{r echo = F}
team_pythag = team_scored_allowed |>
  group_by(team) |>
  summarise(scored = sum(scored),
            allowed = sum(allowed)) |>
  mutate(pythag = round(scored ^ 1.83 / (scored ^ 1.83 + allowed ^ 1.83) * 100, 1))

team_pythag |>
  ggplot(aes(reorder(team, pythag), pythag)) +
  geom_col(aes(fill = team), show.legend = F) +
  coord_flip(ylim = c(0, max(team_pythag$pythag) * 1.05)) +
  scale_fill_manual(values = team_hex) +
  geom_text(aes(label = paste0(pythag, "%")), size = 3, hjust = -0.15) +
  labs(x = NULL, y = "Pythagorean Win Percentage",
       title = glue("Team pythagorean win percentages as of {today_nice}")) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1))
```

___

### Actual vs Pythagorean Win Percentages

```{r echo = F}
team_records |>
  inner_join(team_pythag, by = "team") |>
  ggplot(aes(pythag, win_pct)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  scale_color_manual(values = team_hex) +
  geom_vline(xintercept = 50, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.5) +
  geom_abline(linetype = "dashed", alpha = 0.25) +
  scale_x_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1)) +
  labs(x = "Pythagorean Win Percentage (PWP)",
       y = "Actual Win Percentage",
       title = "Team pythagorean vs. actual win percentage",
       subtitle = "Teams above/below diagonal are worse/better than their record per PWP")
```

___

### Winning and Losing Streaks

```{r echo = F}
team_game_wl = end_games |>
  transmute(date, team = win_team, wl = "win") |>
  bind_rows(end_games |>
  transmute(date, team = lose_team, wl = "loss")) |>
  arrange(team, date)

team_wl_streaks = team_game_wl |>
  group_by(team) |>
  mutate(streak_group = cumsum(wl != lag(wl, default = first(wl))),
         streak = sequence(rle(wl)$lengths),
         streak_type = ifelse(wl == "win", "W", "L"),
         streak_pn = ifelse(streak_type == "W", streak, -streak)) |>
  slice_max(date, n = 1, with_ties = F) |>
  ungroup() |>
  mutate(team_streak = paste0(streak_type, streak))

team_wl_streaks |>
  mutate(pl = ifelse(streak_pn >= 0, team_streak, ""),
         nl = ifelse(streak_pn < 0, team_streak, "")) |>
  ggplot(aes(reorder(team, streak_pn), streak_pn)) +
  geom_col(aes(fill = team), show.legend = F) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  coord_flip() +
  scale_fill_manual(values = team_hex) +
  scale_y_continuous(breaks = seq(-25, 25, by = 1)) +
  labs(x = NULL, y = "Winning/Losing Streak",
       title = glue("Team winning/losing streaks as of {today_nice}"))
```

___

### More to come as the season goes on!

___

### Home and Away Splits

```{r echo = F}
home_wl = end_games |>
  mutate(home_win = ifelse(win_team == home_team, 1, 0),
         away_win = ifelse(win_team == away_team, 1, 0)) |>
  group_by(team = home_team) |>
  summarise(home_wins = sum(home_win),
            home_losses = sum(away_win))

away_wl = end_games |>
  mutate(home_win = ifelse(win_team == home_team, 1, 0),
         away_win = ifelse(win_team == away_team, 1, 0)) |>
  group_by(team = away_team) |>
  summarise(away_wins = sum(away_win),
            away_losses = sum(home_win))

home_wl |>
  full_join(away_wl, by = "team") |>
  mutate(home_wins = coalesce(home_wins, 0),
         home_losses = coalesce(home_losses, 0),
         away_wins = coalesce(away_wins, 0),
         away_losses = coalesce(away_losses, 0),
         home_wp = home_wins / (home_wins + home_losses),
         away_wp = away_wins / (away_wins + away_losses),
         home_wp = coalesce(home_wp, 0),
         away_wp = coalesce(away_wp, 0)) |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(home_wp, away_wp)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  labs(x = "Home Win Percentage", y = "Away Win Percentage",
       title = glue("Home and away win percentages as of {today_nice}")) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), labels = scales::percent)
```

___

### Rolling Window Pythagorean Wins

```{r echo = F}
window_size = 5

team_scored_allowed |>
  arrange(team, date) |>
  mutate(score_roll = rollapply(scored, width = window_size, align = "right", FUN = "sum", fill = NA),
         allow_roll = rollapply(allowed, width = window_size, align = "right", FUN = "sum", fill = NA),
         .by = "team") |>
  filter(!is.na(score_roll) & !is.na(allow_roll)) |>
  mutate(py = (score_roll ^ 2) / (score_roll ^ 2 + allow_roll ^ 2)) |>
  inner_join(teams_info, by = "team") |>
  mutate(window_num = row_number(), .by = "team") |>
  ggplot(aes(window_num, py)) +
  geom_line(aes(col = team), linewidth = 1.5, show.legend = F) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  facet_wrap(vars(abb), scale = "free_x") +
  theme(axis.text = element_blank()) +
  labs(x = "Window number", y = "Pythagorean win percentage",
       title = "Pythagorean win percentages in five-game rolling windows")
```

___

### Rolling Window NPR

```{r echo = F}
team_npr_by_game |>
  transmute(date, team, npr = off_npr + def_npr) |>
  arrange(team, date) |>
  mutate(roll_npr = rollapply(npr, width = window_size, align = "right", FUN = "mean", fill = NA), .by = "team") |>
  filter(!is.na(roll_npr)) |>
  mutate(window_num = row_number(), .by = "team") |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(window_num, roll_npr)) +
  geom_line(aes(col = team), linewidth = 1.5, show.legend = F) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  facet_wrap(vars(abb), scale = "free_x") +
  theme(axis.text = element_blank()) +
  labs(x = "Window number", y = "NPR",
       title = "Team NPR in five-game rolling windows")
```

___

### Records Against Teams with Winning vs Losing Records

```{r echo = F}
team_pct_df = team_records |> select(team, win_pct)

above_below_500_res = end_games |>
  transmute(team = home_team, opp = away_team, win_flg = ifelse(win_team == home_team, 1, 0)) |>
  bind_rows(end_games |>
  transmute(team = away_team, opp = home_team, win_flg = ifelse(win_team == away_team, 1, 0))) |>
  inner_join(team_pct_df, by = c("opp" = "team")) |>
  rename(opp_win_pct = win_pct)

above_below_500_res |>
  mutate(ab500 = ifelse(opp_win_pct >= 50, "above", "below")) |>
  group_by(team, ab500) |>
  summarise(pct = round(mean(win_flg) * 100, 2),
            .groups = "drop") |>
  pivot_wider(id_cols = team, names_from = "ab500", values_from = "pct") |>
  inner_join(teams_info, by = "team") |>
  # this might not be the best way to handle but handles an edge case where
  # the cubs are yet to play a sub .500 opponent as of 04/26/2025
  mutate(above = coalesce(above, below), below = coalesce(below, above)) |>
  ggplot(aes(above, below)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = 50, linetype = "dashed", alpha = 0.25) +
  geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.25) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  labs(x = "Win Percentage vs. Above .500 Teams",
       y = "Win Percentage vs. Below .500 Teams",
       title = glue("Win percentages vs. above/below .500 teams as of {today_nice}")) +
  scale_x_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1))
```

___

### Performance Against Teams with Winning Record

```{r echo = F}
above_500_cnt_pct = above_below_500_res |>
  filter(opp_win_pct >= 50) |>
  group_by(team) |>
  summarise(cnt = n(),
            pct = round(mean(win_flg) * 100)) |>
  inner_join(teams_info, by = "team")

above_500_cnt_pct |>
  ggplot(aes(cnt, pct)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = mean(above_500_cnt_pct$cnt), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 50, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 32) +
  scale_x_continuous(breaks = seq(0, 100, by = 2)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10), labels = scales::percent_format(scale = 1)) +
  labs(x = "Games played against .500 or better teams",
       y = "Win percentage vs. .500 or better teams",
       title = glue("Performance against .500 or better teams as of {today_nice}"))
```

___

### NPR Trends

```{r echo = F}
team_npr_deltas = team_npr_by_game |>
  transmute(date, team, npr = off_npr + def_npr) |>
  arrange(team, date) |>
  mutate(gn = row_number(),
         limit = n() - 10,
         .by = "team") |>
  filter(gn <= limit) |>
  group_by(team) |>
  summarise(npr_prev_x = round(mean(npr), 3)) |>
  inner_join((select(team_npr, team, ovr_npr)), by = "team") |>
  mutate(delta = ovr_npr - npr_prev_x,
         team_label = paste0(team, " (", round(delta, 2), ")"))

team_npr_deltas |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(x = reorder(team_label, ovr_npr), xend = reorder(team_label, ovr_npr),
             y = npr_prev_x, yend = ovr_npr, col = team)) +
  geom_segment(linewidth = 3.5, show.legend = F) +
  geom_text(aes(y = ovr_npr, label = ifelse(delta >= 0, "→", "")),
            size = 4, hjust = -0.2, vjust = 0.25, color = "black", alpha = 0.75) +
  geom_text(aes(y = ovr_npr, label = ifelse(delta < 0, "←", "")),
            size = 4, hjust = 1.2, vjust = 0.25, color = "black", alpha = 0.75) +
  coord_flip() +
  labs(x = NULL, y = "NPR Before/After Last Ten Games",
       title = "Team NPR Trends in Past Ten Games",
       subtitle = "Teams ordered by current NPR; number in parentheses represents change in last ten") +
  scale_y_continuous(breaks = seq(-5, 5, by = 0.25)) +
  scale_color_manual(values = team_hex)
```

___

### Runs Scored and Volatility

```{r echo = F}
team_runs_scored_by_game = end_games |>
  select(date, team = home_team, runs = home_score) |>
  rbind(end_games |>
  select(date, team = away_team, runs = away_score)) |>
  arrange(team, date)

team_runs_avg_sd = team_runs_scored_by_game |>
  group_by(team) |>
  summarise(avg = mean(runs, trim = 0.1),
            sd = sd(runs)) |>
  mutate(lb = avg - sd,
         ub = avg + sd,
         range = ub - lb)

team_runs_avg_sd |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(avg, sd)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = mean(team_runs_avg_sd$avg), linetype = "dashed", alpha = 0.25) +
  geom_hline(yintercept = mean(team_runs_avg_sd$sd), linetype = "dashed", alpha = 0.25) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 32) +
  scale_color_manual(values = team_hex) +
  scale_x_continuous(breaks = seq(0, 10, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 10, by = 0.5)) +
  labs(x = "Avg. Runs Scored", y = "Runs Scored Volatility (SD)",
       title = "Team runs scored vs. volatility",
       subtitle = "Middle 80% of values -- upper & lower 10% (per team) excluded")
```

___

### Performance by Strength of Schedule

```{r echo = F}
team_pythag_small = team_pythag |>
  transmute(team, pythag = pythag / 100)

team_sa = team_scored_allowed |>
  group_by(team) |>
  summarise(scored = sum(scored),
            allowed = sum(allowed))

temp = end_games |>
  select(home_team, away_team, home_score, away_score) |>
  inner_join(team_sa, by = c("home_team" = "team")) |>
  rename(home_scored = scored, home_allowed = allowed) |>
  inner_join(team_sa, by = c("away_team" = "team")) |>
  rename(away_scored = scored, away_allowed = allowed)

team_opp_py = temp |>
  group_by(team = home_team) |>
  summarise(opp_scored = sum(away_scored),
            opp_allowed = sum(away_allowed)) |>
  bind_rows(temp |>
  group_by(team = away_team) |>
  summarise(opp_scored = sum(home_scored),
            opp_allowed = sum(home_allowed))) |>
  group_by(team) |>
  summarise(opp_scored = sum(opp_scored),
            opp_allowed = sum(opp_allowed)) |>
  mutate(opp_py = opp_scored ^ 2 / (opp_scored ^ 2 + opp_allowed ^ 2)) |>
  select(team, opp_py)

team_pythag_small |>
  inner_join(team_opp_py, by = "team") |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(opp_py, pythag)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  labs(x = "Avg. Opponent Pythagorean Win Percentage",
       y = "Pythagorean Win Percentage",
       title = glue("Performance by Strength of Schedule as of {today_nice}")) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.01), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent)
```

___

### One Run Games

```{r echo = F}
one_run_records = end_games |>
  filter(win_score - lose_score == 1) |>
  count(team = win_team) |>
  rename(wins = n) |>
  inner_join(end_games |>
  filter(win_score - lose_score == 1) |>
  count(team = lose_team) |>
  rename(losses = n), by = "team") |>
  mutate(cnt = wins + losses,
         pct = wins / cnt)

one_run_records |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(cnt, pct)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = mean(one_run_records$cnt), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  scale_color_manual(values = team_hex) +
  labs(x = "One-Run Games Played",
       y = "Win Percentage in One-Run Games",
       title = "Performance in One-Run Games") +
  scale_x_continuous(breaks = seq(0, 50, by = 1)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent)
```

___

### Pythagorean Wins in Last Ten

```{r echo = F}
team_scored_allowed |>
  arrange(team, date) |>
  mutate(game_num = row_number(),
         gp = max(game_num),
         .by = "team") |>
  mutate(last7 = ifelse(game_num > gp - 10, 1, 0)) |>
  filter(last7 == 1 & abs(scored - allowed) <= 10) |>
  group_by(team) |>
  summarise(scored = sum(scored),
            allowed = sum(allowed)) |>
  transmute(team,
            py7 = scored ^ 2 / (scored ^ 2 + allowed ^ 2)) |>
  inner_join(team_pythag_small, by = "team") |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(pythag, py7)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  geom_abline(linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  labs(x = "Overall Pythagorean Win Percentage",
       y = "Pythagorean Win Percentage in Last 10 Games",
       title = "Pythagorean Win Percentage: Last 10 Games vs. Full Season") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), labels = scales::percent)
```

___

### Biggest Pythagorean Changes in Last 25

```{r}
team_scored_allowed |>
  arrange(team, date) |>
  mutate(game_num = row_number(),
         games_played = n(),
         .by = "team") |>
  filter(game_num < games_played - 25) |>
  group_by(team) |>
  summarise(scored = sum(scored),
            allowed = sum(allowed)) |>
  transmute(team, py_prev = round(scored ^ 2 / (scored ^ 2 + allowed ^ 2), 3)) |>
  inner_join(team_pythag_small, by = "team") |>
  inner_join(teams_info, by = "team") |>
  mutate(diff = pythag - py_prev) |>
  arrange(desc(diff)) |>
  ggplot(aes(py_prev, pythag)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.25) +
  geom_hline(yintercept = 0.5, linetype = "dashed", alpha = 0.25) +
  # geom_abline(linetype = "solid", col = "red") +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  labs(x = "Pythagorean Wins as of 25 Games Ago",
       y = "Current Pythagorean Wins",
       title = "Pythagorean Movement in Past 25 Games",
       subtitle = "Teams above dashed line have improved in their last 25 relative to other teams") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent)
```

___

```{r}
end_trimmed = end_games |>
  mutate(margin = win_score - lose_score,
         margin_pct = percent_rank(margin)) |>
  slice_min(margin, prop = 0.95)

py_trimmed = end_trimmed |>
  select(team = home_team, scored = home_score, allowed = away_score) |>
  bind_rows(
    end_trimmed |>
      select(team = away_team, scored = away_score, allowed = home_score)
  ) |>
  group_by(team) |>
  summarise(scored = sum(scored),
            allowed = sum(allowed)) |>
  mutate(py_trimmed = scored ^ 2 / (scored ^ 2 + allowed ^ 2),
         lbl = paste0(round(py_trimmed * 100, 1), "%"))

py_trimmed |>
  ggplot(aes(reorder(team, py_trimmed), py_trimmed)) +
  geom_col(aes(fill = team), show.legend = F) +
  coord_flip(ylim = c(0, max(py_trimmed$py_trimmed) * 1.05)) +
  scale_fill_manual(values = team_hex) +
  geom_text(aes(label = lbl), size = 3, hjust = -0.25) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  labs(x = NULL, y = "Adj. Pythagorean Win Percentage",
       title = "Adjusted pythagorean wins",
       subtitle = "All games with a top 5% win margin removed")
```


```{r}
py_adj_scat = team_pythag_small |>
  inner_join(py_trimmed, by = "team") |>
  inner_join(teams_info, by = "team")

py_adj_scat |>
  ggplot(aes(pythag, py_trimmed)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  geom_abline(linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  labs(x = "Actual Pythagorean Win Percentage",
       y = "Adjusted Pythagorean Win Percentage",
       title = "Raw vs. adjusted pythagorean wins",
       subtitle = "Teams below/right of dashed line benefitting from blowout wins")
```

```{r}
help = py_adj_scat |>
  mutate(bb = pythag - py_trimmed,
         real = paste0(round(pythag * 100, 1), "%")) |>
  slice_max(bb, n = 3, with_ties = F) |>
  mutate(desc = as.character(glue("{abb}: {real} real vs. {lbl} adj."))) |>
  pull(desc) |>
  paste0(collapse = "\n")

hurt = py_adj_scat |>
  mutate(bb = pythag - py_trimmed,
         real = paste0(round(pythag * 100, 1), "%")) |>
  slice_min(bb, n = 3, with_ties = F) |>
  mutate(desc = as.character(glue("{abb}: {real} real vs. {lbl} adj."))) |>
  pull(desc) |>
  paste0(collapse = "\n")

py_adj_scat |>
  ggplot(aes(pythag, py_trimmed)) +
  geom_point(aes(col = team), shape = "square", size = 4, show.legend = F) +
  scale_color_manual(values = team_hex) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  geom_abline(linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.05), labels = scales::percent) +
  labs(x = "Actual Pythagorean Win Percentage",
       y = "Adjusted Pythagorean Win Percentage",
       title = "Raw vs. adjusted pythagorean wins",
       subtitle = "Teams below/right of dashed line benefitting from blowout wins") +
  annotate(
    geom = "text",
    label = help,
    x = max(py_adj_scat$pythag) * 0.9,
    y = min(py_adj_scat$py_trimmed) * 1.15,
    size = 3.5,
    fontface = "italic"
  ) +
  annotate(
    geom = "text",
    label = hurt,
    x = min(py_adj_scat$pythag) * 1.25,
    y = max(py_adj_scat$py_trimmed) * 0.95,
    size = 3.5,
    fontface = "italic"
  )

```



```{r}

```



















































